KidsAIR: initial exposure analysis

### 2020-03-19 exposure_initial_analysis.Rmd
exposure analysis updates

---
title: "KidsAIR: initial exposure analysis"
author: "Ethan Walker"
date: "Started 2 Feb 2020, Updated 17 March 2020"
output: pdf_document
---

@@

# SUMs 
```{r}
#### Join PM dataset with full SUMs dataset

pm_join <- read_rds("Output/pm_clean.rds") %>% 
  mutate(sample_datetime = ymd_hms(pm_datetime_new)) %>% 
  separate(sample_datetime, c("sample_date", "sample_time"), sep = " ", remove = FALSE) %>% 
  group_by(area, home) %>% 
  arrange(sample_datetime) %>% 
  mutate(pm_last_datetime = last(sample_datetime),
         pm_first_datetime = first_datetime_new,
         sample_datetime = round_date(sample_datetime, unit = "minute"),
         pm_rolling_20 = rollmean(pm, 20, fill = "extend")) %>% 
  ungroup() %>% 
  arrange(area, home, sample_datetime)

sums_join <- read_rds("Output/sums_clean.rds") %>% 
  mutate(sample_datetime = ymd_hms(datetime_sums)) %>% 
  separate(sample_datetime, c("sample_date", "sample_time"), sep = " ", remove = FALSE) %>% 
  mutate(sample_datetime = round_date(sample_datetime, unit = "minute")) %>% 
  ungroup() %>% 
  arrange(area, home, sample_datetime)

@@

@@

pm_sums_events_joined <- pm_home_mean_date %>% 
  left_join(sums_events_27, by = c("area", "home", "date_sums")) %>% 
  filter(sums_event == 1) %>% 
  arrange(area, home, date_sums) %>% 
  group_by(area, home, mean_pm) %>% 
  summarize(sums_events = n()) %>% 
  #mutate(sums_events = factor(sums_events))
  mutate(sums_events_cat = cut(sums_events, breaks = c(0, 1, 10),
                           labels = c("1", ">1")))

median(pm_sums_events_joined$sums_events)

@@

summary(pm_sums_model)
#exp((0.146291)-1)*100
plot(pm_sums_model, 
     xlab = "fitted values", ylab = "residuals")
qqnorm(residuals(pm_sums_model), main = "QQ Plot")
```

>>>>

---
title: "KidsAIR: initial exposure analysis"
author: "Ethan Walker"
date: "Started 2 Feb 2020, Updated 19 March 2020"
output: pdf_document
---

@@

# SUMs - experimenting with several methods of looking at associations
```{r}
#### Join PM dataset full SUMs dataset by SUMs sampling times

pm_join <- pm_clean %>% 
  mutate(sample_datetime = ymd_hms(pm_datetime_new)) %>% 
  separate(sample_datetime, c("sample_date", "sample_time"), sep = " ", remove = FALSE) %>% 
  group_by(area, home) %>% 
  arrange(sample_datetime) %>% 
  mutate(pm_last_datetime = last(sample_datetime),
         pm_first_datetime = first_datetime_new,
         sample_datetime = round_date(sample_datetime, unit = "minute"),
         pm_rolling_20 = rollmean(pm, 20, fill = "extend")) %>% 
  ungroup() %>% 
  arrange(area, home, sample_datetime)

sums_join <- sums_clean %>% 
  mutate(sample_datetime = ymd_hms(datetime_sums)) %>% 
  separate(sample_datetime, c("sample_date", "sample_time"), sep = " ", remove = FALSE) %>% 
  mutate(sample_datetime = round_date(sample_datetime, unit = "minute")) %>% 
  ungroup() %>% 
  arrange(area, home, sample_datetime)

@@

#### Join PM dataset with SUMs dataset by sampling day

pm_join <- pm_clean %>% 
  filter(winter_id == 1 & treatment == "Placebo") %>% 
  mutate(pm = if_else(pm > 500, 500, pm),
         pm_mean = mean(pm, na.rm = TRUE)) %>% 
  group_by(area, home, sampling_day) %>% 
  mutate(pm_mean_daily = mean(pm, na.rm = TRUE)) %>% 
  mutate(sample_datetime = ymd_hms(pm_datetime_new)) %>% 
  group_by(area, home, sampling_day) %>% 
  arrange(sample_datetime) %>% 
  mutate(pm_last_datetime = last(sample_datetime),
         pm_first_datetime = first(sample_datetime)) %>% 
  ungroup() %>% 
  arrange(area, home, sample_datetime) %>% 
  select(area, home, home_winter_id, dtid, day_of_week, pm_total_observations,
         pm_mean, sampling_day, pm_mean_daily, pm_first_datetime, pm_last_datetime) %>% 
  group_by(area, home) %>% 
  distinct(sampling_day, .keep_all = TRUE) %>% 
  ungroup() %>% 
  arrange(area, home, sampling_day)

sums_join <- sums_clean %>% 
  filter(winter_id == 1 & treatment == "Placebo") %>% 
  select(area, home, winter_id, ibid, treatment, datetime_sums, temp_c_sums,)

pm_sums_joined <- sums_join %>% 
  left_join(pm_join, by = c("area", "home")) %>% 
  arrange(area, home, datetime_sums) %>% 
  mutate(filter_var = if_else(datetime_sums >= pm_first_datetime &
                              datetime_sums <= pm_last_datetime, 1, 0)) %>% 
  filter(filter_var == 1) %>% 
  group_by(area, home, sampling_day) %>% 
  mutate(temp_perc_25 = if_else(temp_c_sums >= 30, 1, 0)) %>% 
  arrange(area, home, sampling_day) %>% 
  mutate(sums_mean_daily = mean(temp_c_sums, na.rm = TRUE),
         sums_mean_daily_5 = sums_mean_daily/5) %>% 
  group_by(area, home, home_winter_id, sampling_day, pm_mean, 
           pm_mean_daily, sums_mean_daily) %>% 
  summarize(temp_perc_25 = (sum(temp_perc_25)/n()*100)) %>% 
  mutate(temp_perc_25_cat = cut(temp_perc_25, breaks = c(0, 10, 200),
                                labels = c("< 50%", "50% +"),
                                 right = FALSE)) %>%
  distinct(pm_mean_daily, .keep_all = TRUE) %>% 
  ungroup() %>% 
  arrange(area, home, sampling_day)

pm_sums_count <- pm_sums_joined %>% 
  count(home)
median(pm_sums_joined$temp_perc_25)
table(pm_sums_joined$temp_perc_25_cat)

cor.test(pm_sums_joined$sums_mean_daily, pm_sums_joined$pm_mean_daily, method = "spearman")
pm_sums_model <- lmer(pm_mean_daily ~ temp_perc_25_cat + (1 | home), data = pm_sums_joined)
pm_sums_model <- lmer(log(pm_mean_daily) ~ temp_perc_25_cat + (1 | home), data = pm_sums_joined)

@@

sums_temp_change <- sums_clean %>% 
  filter(winter_id == 1 & treatment == "Placebo") %>% 
  group_by(area, home, date_sums) %>% 
  mutate(time_diff = (lead(datetime_sums) - datetime_sums)/60) %>% 
  mutate(lead_temp = if_else(time_diff < 20, lead(temp_c_sums, 4), lead(temp_c_sums)),
         lead_temp = if_else(time_diff > 25, 999, lead_temp)) %>% 
  replace_with_na(replace = list(lead_temp = 999)) %>% 
  mutate(temp_diff = lead_temp - temp_c_sums,
         temp_diff_abs = abs(lead_temp - temp_c_sums),
         # change temp in the following line
         temp_diff_check = if_else(temp_diff >= 8, 1, 0),
         heat_event = if_else(temp_diff_check == 1 & lag(temp_diff_check) == 1, 0,
                              temp_diff_check),
         heat_event = if_else(is.na(temp_diff_check), 1,
                              heat_event)) %>% 
  ungroup() %>% 
  mutate(date_sums = as.character(date_sums)) %>% 
  group_by(area, home, date_sums) %>% 
  summarize(sums_events = sum(heat_event, na.rm = TRUE)) %>% 
  ungroup() %>% 
  arrange(area, home, date_sums)

pm_sums_events_joined <- pm_home_mean_date %>% 
  left_join(sums_temp_change, by = c("area", "home", "date_sums")) %>% 
  arrange(area, home, date_sums) %>% 
  #mutate(sums_events = factor(sums_events))
  mutate(sums_events_cat = cut(sums_events, breaks = c(0, 2, 20),
                           labels = c("<2", "2+"),
                                 right = FALSE))

median(pm_sums_events_joined$sums_events, na.rm = TRUE)
table(pm_sums_events_joined$sums_events_cat)

@@

summary(pm_sums_model)
#exp((0.146291)-1)*100
plot(pm_sums_model, 
     xlab = "fitted values", ylab = "residuals")
qqnorm(residuals(pm_sums_model), main = "QQ Plot")

pm_sums_plot <- pm_sums_events_joined %>% 
  filter(!is.na(sums_events)) %>% 
  ggplot() + 
    geom_boxplot(aes(sums_events_cat, log(mean_pm))) +
    theme_minimal()
pm_sums_plot

```

# Create new dataset with mean PM/SUMs by PM sampling day; join with covariates
```{r}
# Prep PM data to join with SUMs data by sampling day
pm_join <- pm_clean %>% 
  filter(winter_id == 1 & treatment == "Placebo") %>% 
  group_by(area, home) %>% 
  mutate(pm = if_else(pm > 500, 500, pm),
         pm_mean = mean(pm, na.rm = TRUE)) %>% 
  group_by(area, home, sampling_day) %>% 
  mutate(pm_mean_daily = mean(pm, na.rm = TRUE)) %>% 
  mutate(sample_datetime = ymd_hms(pm_datetime_new)) %>% 
  group_by(area, home, sampling_day) %>% 
  arrange(sample_datetime) %>% 
  mutate(pm_last_datetime = last(sample_datetime),
         pm_first_datetime = first(sample_datetime)) %>% 
  ungroup() %>% 
  arrange(area, home, sample_datetime) %>% 
  select(area, home, home_winter_id, dtid, day_of_week, pm_total_observations,
         pm_mean, sampling_day, pm_mean_daily, pm_first_datetime, pm_last_datetime) %>% 
  group_by(area, home) %>% 
  distinct(sampling_day, .keep_all = TRUE) %>% 
  ungroup() %>% 
  arrange(area, home, sampling_day)

# Prep SUMs data to join with PM data
sums_join <- sums_clean %>% 
  filter(winter_id == 1 & treatment == "Placebo") %>% 
  select(area, home, winter_id, ibid, treatment, datetime_sums, temp_c_sums,)

# Join PM/SUMs data and filter and arrange by sampling day
pm_sums_joined <- sums_join %>% 
  left_join(pm_join, by = c("area", "home")) %>% 
  arrange(area, home, datetime_sums) %>% 
  mutate(filter_var = if_else(datetime_sums >= pm_first_datetime &
                              datetime_sums <= pm_last_datetime, 1, 0)) %>% 
  filter(filter_var == 1) %>% 
  group_by(area, home, sampling_day) %>% 
  mutate(temp_perc_30 = if_else(temp_c_sums >= 30, 1, 0)) %>% 
  arrange(area, home, sampling_day) %>% 
  mutate(sums_mean_daily = mean(temp_c_sums, na.rm = TRUE),
         sums_mean_daily_5 = sums_mean_daily/5) %>% 
  group_by(area, home, home_winter_id, sampling_day, 
           sums_mean_daily, sums_mean_daily_5) %>% 
  summarize(temp_perc_30 = (sum(temp_perc_30)/n()*100)) %>% 
  mutate(temp_perc_30_cat = cut(temp_perc_30, breaks = c(0, 10, 200),
                                labels = c("< 10%", "10% +"),
                                 right = FALSE)) %>%
  distinct(pm_mean_daily, .keep_all = TRUE) %>% 
  ungroup() %>% 
  arrange(area, home, sampling_day)

# Join SUMs data with original PM data, by sampling day
pm_sums_joined2 <- pm_sums_joined %>% 
  right_join(pm_join, by = c("area", "home", "home_winter_id", "sampling_day")) %>% 
  mutate(sampling_day = as.factor(sampling_day)) %>% 
  ungroup() %>% 
  arrange(area, home, home_winter_id, sampling_day)

# Join PM/SUMs data with other covariate datasets
analysis_data1 <- home_act_clean %>% 
  filter(winter_id == 1 & treatment == "Placebo") %>% 
  mutate(sampling_day = home_act_day) %>% 
  select(-home_act_day, -home_act_comments, -home_id) %>% 
  left_join(pm_sums_joined2, by = c("area", "home", "home_winter_id", "sampling_day")) %>% 
  ungroup() %>% 
  arrange(area, home, home_winter_id, sampling_day)

analysis_data2 <- demographics_clean %>% 
  filter(winter_id == 1 & treatment == "Placebo") %>% 
  select(area, home, home_winter_id, gender_parent:residents_under_five) %>% 
  distinct(home, .keep_all = TRUE) %>% 
  left_join(analysis_data1, by = c("area", "home", "home_winter_id")) %>% 
  filter(!is.na(gender_parent)) %>% 
  mutate(income_3level = fct_collapse(income,
                               "Less than $20,000" = "Less than $20,000",
                               "$20,000-$49,999" = 
                                 c("$20,000 to $29,999", "$30,000 to $39,999",
                                   "$40,000 to $49,999"),
                                "$50,000+" = 
                                 c("$50,000 to $74,999", "$75,000 to $99,999",
                                   "$100,000 or more")),
         income_3level = factor(income_3level, levels = c("Less than $20,000", 
                                            "$20,000-$49,999", 
                                            "$50,000+")), 
         education_3level = fct_collapse(education,
                                  "High school or less" = 
                                    c("Less than high school", 
                                      "High school diploma or GED"),
                                  "Some college" = "Some college",
                                  "College degree" = "College degree"),
         education_3level = factor(education_3level, levels = c("High school or less", 
                                            "Some college", 
                                            "College degree")), 
         race_parent_2level = fct_collapse(race_parent,
                                    "White" = "White",
                                    "Non-white" = 
                                      c("American Indian/Alaskan Native",
                                        "Asian", "More than one race")),
         race_parent_2level = factor(race_parent_2level, levels = c("White", "Non-white")), 
         total_residents_2level = fct_collapse(total_residents,
                                    "<5" = c("1", "2", "3", "4"),
                                    "5+" = c("5", "6", "7", "8", "9", "10",
                                             "11", "12", "13", "14", "15")),
         total_residents_2level = factor(total_residents_2level, levels = c("<5", "5+")),
         residents_under5_2level = fct_collapse(residents_under_five,
                                    "1" = c("0", "1"),
                                    "2+" = c("2", "3", "4", "5")),
         residents_under5_2level = factor(residents_under5_2level, levels = c("1", "2+"))) %>% 
  ungroup() %>% 
  arrange(area, home, home_winter_id, sampling_day)

analysis_data3 <- home_char_clean %>% 
  filter(winter_id == 1 & treatment == "Placebo") %>% 
  select(area, home, home_winter_id, residents_smoke:home_insert) %>% 
  distinct(home, .keep_all = TRUE) %>% 
  right_join(analysis_data2, by = c("area", "home", "home_winter_id")) %>% 
  mutate(home_type_2level = fct_collapse(home_type,
                                  "House" = "House",
                                  "Other" = c("Duplex/Apt", "Mobile Home", "Other")),
         home_type_2level = factor(home_type_2level,
                            levels = c("House", "Other")),
         home_year_built_2level = cut(home_year_built, breaks = c(0, 1985, 2020),
                                 labels = c("<1985", "1985+"),
                                 right = FALSE),
         stove_age = factor(stove_age,
                         levels = c("0-5 years old", "6-10 years old",
                                    "11-15 years old", "16 + years old",
                                    "Do not know")),
         stove_age_4level = fct_collapse(stove_age,
                                  "0-5 years old" = "0-5 years old",
                                  "6-15 years old" = c("6-10 years old",
                                                       "11-15 years old"),
                                  "16 + years old" = "16 + years old",
                                  "Do not know" = "Do not know"),
         stove_cert = factor(stove_cert,
                             levels = c("Yes", "No", "Do not know")),
         home_sqft_2level = cut(home_sqft, breaks = c(0, 1400, 5000),
                         labels = c("<1400", "1400+"),
                         right = FALSE),
         home_floors_2level = cut(home_floors, breaks = c(0, 2, 10),
                                 labels = c("1", "2+"),
                                 right = FALSE),
         home_bedrooms_2level = cut(home_bedrooms, breaks = c(0, 3, 50),
                                 labels = c("<3", "3+"),
                                 right = FALSE),
         home_windows_2level = cut(home_windows, breaks = c(0, 10, 100),
                                labels = c("<10", "10+"),
                                 right = FALSE),
         home_pets_2level = cut(home_pets, breaks = c(0, 2, 50),
                                labels = c("<2", "2+"),
                                 right = FALSE),
         home_furry_2level = cut(home_furry, breaks = c(0, 2, 50),
                                labels = c("<2", "2+"),
                                 right = FALSE),
         secondary_heat_3level = fct_collapse(secondary_heat,
                                       "Electric" = "Electrical",
                                       "Wood stove" = "Wood stove",
                                       "Other" = c("Natural gas furnace",
                                                   "Oil", "Other", "Propane")),
         chimney_clean = factor(chimney_clean,
                         levels = c("Less than 6 months ago", "6-12 months ago",
                                    "12-18 months ago", "More than 18 months ago")),
         chimney_clean_3level = fct_collapse(chimney_clean,
                                  "Less than 6 months ago" = "Less than 6 months ago",
                                  "6-12 months ago" = "6-12 months ago",
                                  "12 + months ago" = c("12-18 months ago",
                                                        "More than 18 months ago")),
         home_damp = factor(home_damp, levels = c("Yes", "No", "Do not know")),
         home_mold = factor(home_mold, levels = c("Yes", "No", "Do not know")),
         home_fireplace = factor(home_fireplace, levels = c("Yes", "No", "Do not know")),
         home_furnace = factor(home_furnace, levels = c("Yes", "No", "Do not know")),
         home_insert = factor(home_insert, levels = c("Yes", "No", "Do not know"))) %>% 
  ungroup() %>% 
  arrange(area, home, home_winter_id, sampling_day)

analysis_data4 <- indoor_temp_rh_clean %>% 
  filter(winter_id == 1 & treatment == "Placebo") %>% 
  select(area, home, home_winter_id, rh_indoor_current:temp_indoor_min) %>% 
  distinct(home, .keep_all = TRUE) %>% 
  right_join(analysis_data3, by = c("area", "home", "home_winter_id")) %>% 
  mutate(temp_max_5 = temp_indoor_max/5,
         temp_min_5 = temp_indoor_min/5,
         rh_max_5 = rh_indoor_max/5,
         rh_min_5 = rh_indoor_min/5,
         temp_max_2level = cut(temp_indoor_max, breaks = c(0, 30, 100),
                                labels = c("<30", "30+"),
                                 right = FALSE),
         temp_min_2level = cut(temp_indoor_min, breaks = c(0, 16.5, 100),
                                labels = c("<16.5", "16.5+"),
                                 right = FALSE),
         rh_max_2level = cut(rh_indoor_max, breaks = c(0, 40, 200),
                                labels = c("<40", "40+"),
                                 right = FALSE),
         rh_min_2level = cut(rh_indoor_min, breaks = c(0, 20, 200),
                                labels = c("<20", "20+"),
                                 right = FALSE)) %>% 
  ungroup() %>% 
  arrange(area, home, home_winter_id, sampling_day)

analysis_data5 <- moisture_filtered_time %>% 
  filter(winter_id == 1 & treatment == "Placebo") %>% 
  select(area, home, home_winter_id, moisture_date, moisture_ave, moisture_split) %>% 
  distinct(home, .keep_all = TRUE) %>% 
  right_join(analysis_data4, by = c("area", "home", "home_winter_id")) %>% 
  mutate(moisture_ave_5 = moisture_ave/5,
         moisture_2level = cut(moisture_ave, breaks = c(0, 10, 100),
                                labels = c("<10", "10+"),
                                 right = FALSE)) %>% 
  ungroup() %>% 
  arrange(area, home, home_winter_id, sampling_day)

analysis_data6 <- stove_grades_clean %>% 
  arrange(area, home, winter_id) %>% 
  group_by(home) %>% 
  mutate(stove_grade = if_else(is.na(stove_grade), lead(stove_grade), stove_grade)) %>% 
  filter(winter_id == 1 & treatment == "Placebo") %>% 
  select(area, home, home_winter_id, stove_grade) %>% 
  distinct(home, .keep_all = TRUE) %>% 
  right_join(analysis_data5, by = c("area", "home", "home_winter_id")) %>% 
  mutate(stove_grade_3level = fct_collapse(stove_grade,
                                    "A" = "A",
                                    "C" = "C",
                                    "D or F" = c("D", "F"))) %>% 
  ungroup() %>% 
  arrange(area, home, home_winter_id, sampling_day)

analysis_data7 <- stove_use_clean %>% 
  filter(winter_id == 1 & treatment == "Placebo") %>% 
  select(area, home, home_winter_id, burn_level:wood_cords) %>% 
  distinct(home, .keep_all = TRUE) %>% 
  right_join(analysis_data6, by = c("area", "home", "home_winter_id")) %>% 
  mutate(burn_level_3level = fct_collapse(burn_level,
                                   "No/Light burning" = c("No burning", 
                                                          "Light burning"),
                                   "Ave burning" = "Average burning",
                                   "Heavy burning" = "Heavy burning"),
         burn_level_3level = factor(burn_level_3level,
                         levels = c("No/Light burning", "Ave burning", "Heavy burning")),
         wood_collect_2level = fct_collapse(wood_collect,
                                   "<3 months" = c("< 1 week", "1 week to 1 mo",
                                                   "1 mo to 3 mo"),
                                   "3+ months" = c("3 mo to 6 mo", 
                                                   "6 mo to 1 year", "> 1 yr")),
         wood_collect_2level = factor(wood_collect_2level,
                         levels = c("<3 months", "3+ months")),
         wood_collect_method_2level = fct_collapse(wood_collect_method,
                                   "Harvest yourself" = "Harvest yourself",
                                   "Purchase/other" = c("Mix", "Other", 
                                                        "Purchase it")),
         wood_cords = if_else(wood_cords > 10, 10, wood_cords),
         wood_cords_2level = cut(wood_cords, breaks = c(0, 5, 50),
                                labels = c("<5", "5+"),
                                 right = FALSE)) %>% 
  ungroup() %>% 
  arrange(area, home, home_winter_id, sampling_day)

#write_rds(analysis_data7, "Output/exposure_analysis_data.rds")
#write_csv(analysis_data7, "Output/exposure_analysis_data.csv")
```
